---
title: "Repsc vignette"
author: "David Brocks"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Repsc}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(eval = FALSE)
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

# Workflow human 10x scRNA-seq dataset (5')
In this tutorial, we are going to utilize 5' scRNA-seq data on epigenetically de-repressed cancer cell lines to quantify transposable element (TE) expression levels at single-cell and locus resolution. Following the workflow, you'll learn the specifics of Repsc to adapt it to your single-cell dataset.


## Getting started

We start the workflow by loading Repsc and the human hg38 _BSgenome_ object into our R environment:
```{r, eval = TRUE, echo = TRUE, warning = FALSE}
Sys.time()
devtools::load_all('/net/mraid14/export/tgdata/users/davidbr/src/Repsc/')
devtools::load_all('/net/mraid14/export/tgdata/users/davidbr/src/Reputils/')
devtools::load_all('/net/mraid14/export/tgdata/users/davidbr/src/Repdata/')

library(BSgenome.Mmusculus.UCSC.mm10)
```

## Create scSet

We then import our gene and TE annotation files as [GRanges objects](https://bioconductor.org/packages/release/bioc/vignettes/GenomicRanges/inst/doc/GenomicRangesIntroduction.html) followed by Repsc-specific curation and formatting using the `curateGenes` and `curateTEs` functions.
```{r, eval = TRUE, echo = TRUE, warning = FALSE}
# path to Gencode gtf file (provided)
gene_path <- system.file(package = 'Repdata', 
                        'extdata', 
                        'mm10',
                        'genes',
                        'gencode.vM22.annotation.gtf.gz')

# path to RepeatMasker mm10 repeat annotation (provided)
rmsk_path <- system.file(package = 'Repdata', 
                         'extdata', 
                         'mm10',
                         'tes',
                         'mm10.fa.out.gz')
                         
# creating the scSet
sc <- createScSet(genome   = Mmusculus,
                  protocol = 'fiveprime',
                  tes      = rmsk_path,
                  genes    = gene_path)
```

## Create the input data.frame

After we have imported and curated the data, we can procede to generate the actual read/UMI count matrix using the `compCounts` function.
 
```{r, eval = TRUE, echo = TRUE, warning = FALSE}
# path to bam files containing mapped reads                         
bam_paths <-  dir('/net/mraid14/export/data/users/davidbr/proj/eseb/data/', 
                  recursive = TRUE, 
                  pattern = '_deduplicated_chr[0-9, X, Y, MT]+.bam', 
                  full.names = TRUE)
                  
hdf5_paths <- dir('/net/mraid14/export/data/users/davidbr/proj/eseb/data/',
                  recursive = TRUE,
                  pattern = 'filtered_gene_bc_matrices_h5.h5',
                  full.names = TRUE)

# create a data.frame specifying import parameters                 
input_df    <- data.frame(paths   = bam_paths,
                          hdf5    = rep(hdf5_paths, each = 22),
                          paired  = TRUE,       # use FALSE for single-end libraries
                          mate    = 'first',    # only imports the first mate of properly aligned read pairs, set to NA when using single-end libraries
                          barcode = 'CB',       # 10x barcode included in BAM flag
                          chunk   = ntile(bam_paths, 40),
                          meta    = gsub('/', '', substring(bam_paths, 56, 61)),
                          stringsAsFactors = FALSE)
                        
head(input_df)                      
```

```{r, eval = TRUE, echo = TRUE, warning = FALSE}
sc <- addCounts(sc,
                bams     = input_df,
                bin_size = 25,
                msa_dir  = NULL,
                use_gcluster = TRUE)
```

## Call cells

To distinguish real cells from empty droplets, we utilize the `emptyDrops` function from the [DropletUtils](https://rdrr.io/github/MarioniLab/DropletUtils/) package[1].

```{r, eval = TRUE, echo = TRUE, warning = FALSE, fig.width = 7, fig.height = 8, fig.align='center'}
plotCells(sc)
```

## Mapping
```{r, eval = TRUE, echo = TRUE, warning = FALSE, fig.width = 7, fig.height = 7, fig.align='center'}
plotMapping(sc)
```

# Call peaks
```{r, eval = TRUE, echo = TRUE, warning = FALSE, fig.width = 7, fig.height = 12, fig.align='center'}
sc <- selectPeaks(sc)
plotPeaks(sc)
```

# Feature selection
```{r, eval = TRUE, echo = TRUE, warning = FALSE, fig.width = 7, fig.height = 7, fig.align='center'}
sc <- selectFeatures(sc)
plotFeatures(sc)
```


      
# References
[1]   

# Session information
```{r}
sessionInfo()
```          